---
title: "IDS investigation worksheet"
author: "by Team31: Finn, Tom, Jon, Scott & Roman"
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, message = FALSE}
library(tidyverse)
library(tidymodels)
# Add any other libraries here



```


```{r load-data}
# setwd("~/IDS/GroupProject/data/census+income")
adult_data <- read.csv("data/adult.data")

#Changing the variable names for adult data
colnames(adult_data) <- c("age", "workclass", "id", "education", "education-num", "marital-status", "occupation", "relationship", "race_old", "sex", "capital_gain", "capital_loss", "hours_per_week", "native-country", "class")

adult_data <- adult_data %>% 
  select(-capital_gain, -capital_loss)

#Here we have transformed the race variable to make it binary depending on if the person is white or not (1 for white 0 for other)
is_white <- function(x){ifelse(x == " White", 1, 0)}

adult_data <- adult_data %>% mutate(race = is_white(race_old))
adult_data <- adult_data %>% select(-race_old)


#Here we have created a new "has degree" variable which is a binary value taking on 1 if the person has a degree
has_degree <- function(x){ifelse(x == " Bachelors" | x == " Masters" | x == " Doctorate", 1, 0)}

adult_data <- adult_data %>% mutate(has_degree = has_degree(education))


view(adult_data)

```

```{r plot degree vs class}
#some questions about the data: income inequality between different races/genders/nationalities ; class based on type of degree ; correlation between class and age; correlation between race/gender and age E.G do certain races/genders live longer than others.
education_counts <- adult_data %>%
  group_by(education) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

adult_data$education <- adult_data$education %>% factor(levels = education_counts$education)

adult_data %>% ggplot() + geom_bar(mapping = aes(x = education, fill = class), position = "dodge") + theme(axis.text.x = element_text(angle = 45, hjust = 1))

#from this graph it looks as if bachelors degrees have the most people in upper class, however this does not take into account the fact that more people have bachelos in comparison to master, phd, etc..
```


```{r plot sex vs class as percentage}
percentage_sex_data <- adult_data %>%
  group_by(sex, class) %>%
  summarize(count = n()) %>%
  mutate(percentage = count / sum(count) * 100)


percentage_sex <- ggplot(percentage_sex_data, aes(x = sex, y = percentage, fill = class)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), vjust = -0.5, position = position_dodge(width = 0.9)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(percentage_sex)
```
```{r program to determine income}
#Here we have created a program which predicts a persons class based on the other variable.
#The program is made using conditional probabilities so we can compare this model to that of a linear regression once we make a linear regression model.

conditional_probs <- with(adult_data, prop.table(table(class, race, sex, education), margin = 1))

#please put a space before enetering each value and also use capitals at start
new_individual <- data.frame(race = " White", sex = " Male", education = " Bachelors")

# Calculate the conditional probabilities for both classes
conditional_probs_above_50K <- conditional_probs[" >50K", new_individual$race, new_individual$sex, new_individual$education]
conditional_probs_below_50K <- conditional_probs[" <=50K", new_individual$race, new_individual$sex, new_individual$education]

# Compare the conditional probabilities and classify the individual
if (conditional_probs_above_50K > conditional_probs_below_50K) {
  result = paste("class is >50K With probability", conditional_probs_above_50K, sep = " ")
  print(result)

} else {
  result = paste("Class is <= 50K With probability", conditional_probs_below_50K, sep = " ")
  print(result)
}
```

``` {r linear regression using sex, rae degree as predictors}

adult_data$class <- factor(adult_data$class)

class_fit <- logistic_reg() %>%
  set_engine("glm") %>%
  fit(class ~ race + sex + has_degree, data = adult_data, family = "binomial")

tidy(class_fit)

```


``` {r age vs class plot }
adult_data %>% ggplot() + geom_bar(mapping = aes(x = age, colour = class))

mean_age <- adult_data %>% group_by(class) %>% summarize(mean_age = mean(age))
mean_age

```

since sex is such a good predictor of class (it has a p-value of 3.37e-290), we suspect that maybe sex is related to one of the other predictors. clearly there are no relationships between sex and race - hence we should calculate the probabiltiy of having a degree given that a person is male vs female
``` {r creating a logistic regression }

```
hence p = 24% for males
      p = 21% for females 
      
Although this is somewhat of a large discrepancy it does not fully explain the extremely low p-value for sex as a predictor of wealth 

hence we then decided to predict wealth based on only sex alone

``` {r creating a logistic regression }

adult_data$class <- factor(adult_data$class)

class_sex_fit <- logistic_reg() %>%
  set_engine("glm") %>%
  fit(class ~ sex, data = adult_data, family = "binomial")

tidy(class_sex_fit)
```

so for this model males have a 23% chance of being upper class
however females have a 21% chance
being males increases your chance of being upper class by a factor of 1.189




it may not come as a suprise that more people are rich etc... and older people are rich...

```{r age vs hours per week linear regression}

ggplot(data = adult_data, aes(x = age, y = hours_per_week)) +
  geom_point() +
  geom_smooth(method = "lm", 
              se = FALSE) +
  labs(
    title = "Age vs hours worked per week",
    x = "age (years)",
    y = "hours-per-week"
  )

linear_reg() %>%
  set_engine("lm") %>%
  fit(hours_per_week ~ age, data = adult_data) %>%
  tidy()

```
so hours_per_week = 38 + 0.062(age)

so older people are more likely to have a greater income however they also tend to work more hours which may have some correlation. We will now take a look at this. Note that for every year older a person is the number of hour per week increases by 0.062 hours.

```{r logistic regression for hours-per-week vs income class}

adult_data$class <- factor(adult_data$class)

class_fit <- logistic_reg() %>%
  set_engine("glm") %>%
  fit(class ~ hours_per_week, data = adult_data, family = "binomial")

tidy(class_fit)

```
so as a persons hours worked per week increased by one hour the probability of a person being upper-class increases by a factor of 4.6%

Hence we may

list of things to do for presentation:
fix current slide of gender vs class
make a slide for each of the graphs and add some text explaining it

create plot of number of hours worked vs gender

```{r}
adult_data %>% ggplot() + geom_bar(mapping = aes(x = race, fill = class), position = "dodge") + labs( x = "Not White                                                        White")
prop_white <- adult_data %>% filter(race == 1) %>% summarise(prop_white = sum(class == " >50K")/sum(class == " <=50K"))
prop_white

prop_not_white <- adult_data %>% filter(race == 0) %>% summarise(prop_not_white = sum(class == " >50K")/sum(class == " <=50K"))
prop_not_white
```

```{r class vs hour-per-week, echo = FALSE}
adult_data$class <- factor(adult_data$class)

class_fit <- logistic_reg() %>%
  set_engine("glm") %>%
  fit(class ~ hours_per_week, data = adult_data, family = "binomial")

tidy(class_fit)
```

```{r class-vs-age, echo = FALSE}
adult_data$class <- factor(adult_data$class)

class_fit <- logistic_reg() %>%
  set_engine("glm") %>%
  fit(class ~ age, data = adult_data, family = "binomial")

tidy(class_fit)
```