---
title: "Interpreting census data"
subtitle: "Investigating the relationships between income and personal variables"
author: "The Best Team <br> Finn, Tom, Jon, Scott, Roman"
institute: "University of Edinburgh"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      
---

```{r install-package, include = FALSE, eval = FALSE}
# Copy and paste the following code into your console to download and install
# the `xaringan` package that contains all of the code which allows you 
# to create presentation slides in Rmarkdown
install.packages('xaringan')
```


```{r load-packages, include = FALSE}
# Add any additional packages you need to this chunk
library(tidyverse)
library(tidymodels)
library(palmerpenguins)
library(knitr)
library(xaringanthemer)
```

```{r setup, include=FALSE}
# For better figure resolution
knitr::opts_chunk$set(fig.retina = 3, dpi = 300, fig.width = 6, fig.asp = 0.618, out.width = "80%")
```

```{r load-data, include=FALSE}

adult_data <- read.csv("data/adult.data")


```

```{r include=FALSE}

#Background image
style_xaringan(
  title_slide_background_image = "img/background.jpg"
)
```

class: center, middle

## Research into which demographics have the highest proportion of upper-class citizens

---
class: center, middle

## About Our Dataset

---
## Adult Data

Our main data set is called "Adult data" and contains details of over 3300 people from a Census taken in the US in 1994. Note that all of those participating in the census are over 16 years old. The key variables used in our data set are as follows:

--

- Age

--

- Education

--

- Sex

--

- Hour worked per week

--

- Race

--

- Income ( >50K or <50K)

---
class: center, middle

## Data Wrangling

---
##Wrangling our dataset

Our data set did not contain any column names so our first call of action was to name the variables.

```{r}
#Changing the variable names for adult data
colnames(adult_data) <- c("age", "workclass", "id", "education", "education-num", "marital-status",
                          "occupation", "relationship", "race_old", "sex", "capital_gain",
                          "capital_loss", "hours_per_week", "native-country", "class")
```

---
##Creating binary variables

- We creates a new variable called "has_degree" which takes on 1 if a person has bachelors, masters or PHD and 0 otherwise

- similarly we mutated the race variable such that it takes on 1 if a person is white and 0 otherwise. Although this means we lose some information it allows us to simplify our predictions and still lets us see discrimination towards minorities.

```{r}
#Here we have transformed the race variable to make it binary depending on if the person is white or not (1 for white 0 for other)
is_white <- function(x){ifelse(x == " White", 1, 0)}

is_married <- function(x){ifelse(x == " Married-civ-spouse", 1,0)}
adult_data <- adult_data %>% mutate(is_married = is_married(`marital-status`) )

adult_data <- adult_data %>% mutate(race = is_white(race_old))

has_degree <- function(x){ifelse(x == " Bachelors" | x == " Masters" | x == " Doctorate", 1, 0)}
adult_data <- adult_data %>% mutate(has_degree = has_degree(education))

adult_data <- adult_data %>% select(-capital_gain, -capital_loss, -race_old)
```

---
class: center, middle
#Analysis 1 
##How age and work hours relate to income
---
##How are age and class related

Observe that the overall distributions are similar however there are much more lower-class people and they generally have a younger age:

- We used position = "identity" to have the two plots overlap each other
- Mean age for upper class is 44
- Mean age for lower-class is 37


```{r warning=FALSE, out.width="50%", fig.width=4, echo=FALSE}
adult_data %>% ggplot() + geom_bar(mapping = aes(x = age, colour = class), position = "identity")
```
---
##So why are upper-class people older?
One such possibility is that people generally tend to work more hours as they get older hence they may make more money for working more hours. 

```{r warning=FALSE, out.width="50%", fig.width=4, echo=FALSE}
adult_data %>% ggplot() + geom_smooth(mapping = aes(x = age, y = hours_per_week))
```


---
##Additional evidence

So clearly we can see that as age increases so too does the number of hours worked
namely, we have:

$\text{hours_per_week} = 38 + 0.0622 \times \text{age}$



```{r warning=FALSE, out.width="50%", fig.width=4, echo=FALSE}

ggplot(data = adult_data, aes(x = age, y = hours_per_week)) +
  geom_point() +
  geom_smooth(method = "lm", 
              se = FALSE) +
  labs(
    title = "Age vs hours worked per week",
    x = "age (years)",
    y = "hours-per-week"
  )

linear_reg() %>%
  set_engine("lm") %>%
  fit(hours_per_week ~ age, data = adult_data) %>%
  tidy()

```

---
##Looking at residuals

```{r, echo = FALSE}
hours_age_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(hours_per_week ~ age, data = adult_data)
  

hours_age_fit_aug <- augment(hours_age_fit$fit)
ggplot(hours_age_fit_aug, mapping = aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "gray", lty = "dashed") +
  labs(x = "Predicted hours worked", y = "Residuals")
```
---
##But does working more relate to earning more?
A logistic regression was made for the class variable using hours_per_week as the predictor:
```{r class vs hour-per-week, echo = FALSE}
adult_data$class <- factor(adult_data$class)

class_fit <- logistic_reg() %>%
  set_engine("glm") %>%
  fit(class ~ hours_per_week, data = adult_data, family = "binomial")

tidy(class_fit)
```
-  Hence some basic calculations show that for every year older a person is their probability of being predicted as upper class increases by roughly 0.2%
---
##Conclusive thoughts
 
 So in this first analysis we came to the conclusion that older people tend to be financially better off. We attempt to give an explanation to this using data analysis in the following way:
 
--
 
 - Older people tend to work more hours
 
--

 - People who work more hours tend to earn more
 
--
 
 - Therefore older people have greater chance of earning more


---
class: center, middle
##Analysis 2
##Looking at how sex, race and degree relate to income class
---
#The relationship between sex and class
```{r warning=FALSE, out.width="50%", fig.width=6, echo=FALSE}
# see how I changed out.width and fig.width from defaults
# to make the figure bigger
percentage_sex_data <- adult_data %>%
  group_by(sex, class) %>%
  summarize(count = n()) %>%
  mutate(percentage = count / sum(count) * 100)

percentage_sex <- ggplot(percentage_sex_data, aes(x = sex, y = percentage, fill = class)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), vjust = -0.5, position = position_dodge(width = 0.9)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

percentage_sex ##create some more pages with graphs and explaning them
```

---
##But what causes such a discrepency between the class of each sex?
One possible explanation is again the possibility that one sex may have tended to work more hours than another (keep in mind that this data is from 1994 where the different attidues towards each sex were much more previlent.)
.pull-left[- Mean hours worked per week for males in 42.4
- Mean hours worked per week for females is 36.4
- notice that the red line is higher on the left and the blue takes over on the right


hence from before since people who work more tend to earn more this may be one key component of the many different contributers to this class inequity between the two sex's.]
``` {r,warning=FALSE, out.width="50%", fig.width=6, echo=FALSE}
adult_data %>% ggplot() + geom_density(mapping = aes(x = hours_per_week, colour = sex))

```

---
##The relationship between education and class

```{r warning=FALSE, echo=FALSE}
education_counts <- adult_data %>%
  group_by(education) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

adult_data$education <- adult_data$education %>% factor(levels = education_counts$education)

adult_data %>% ggplot() + geom_bar(mapping = aes(x = education, fill = class), position = "dodge") + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
---
## A closer look into those with and without a degree
```{r, echo = FALSE}
adult_data %>% ggplot() + geom_bar(mapping = aes(x = has_degree, fill = class), position = "dodge") + labs( x = "No Degree                           Degree")

#Note that 47% of people with degrees are upper-class whereas only 17% of people without degrees are upper-class
```
---
##The relationship between race and class
```{r warning=FALSE, echo=FALSE}
adult_data %>% ggplot() + geom_bar(mapping = aes(x = race, fill = class), position = "dodge") + labs( x = "Not White                                                        White")

# 34% of white people are upper class
# 18% of not white people are upper class.


```

---
#Logistic regression results

After careful consideration we came to conclusion that sex, race and university degrees were the most interesting predictors of a persons income and also from the plots above there seems to be a clear disparity within them. Therefore, in an attempt to keep our model as simple as possible, our logistic regression uses only sex, race and degree as predictors. This resulted in the following table:

```{r logistics-table, echo = FALSE}
adult_data$class <- factor(adult_data$class)

class_fit <- logistic_reg() %>%
  set_engine("glm") %>%
  fit(class ~ race + sex + has_degree, data = adult_data, family = "binomial")

tidy(class_fit)
```
note that we transformed the data in such a way that race takes on the value 1 if the person is white and 0 otherwise. This is because we wanted to see specifically how being in a minority may affect people and it also allows the data to be further simplified leading to better predictions.


---
#Creating our prediction model

By using the data from the previous slide we arrive at the following logit function model:

 $\log\left(\frac{p}{1 - p}\right) = -2.96 + 0.494 \times \text{race} + 1.3 \times \text{sex} + 1.45 \times \text{degree}$

<br>

--
Which means that:

$p = \frac{1}{1 + e^{2.96 - 0.494 \times \text{race} - 1.3 \times \text{sex} - 1.45 \times \text{degree}}}$

--

We then turned this data into a table to look at the results. Notice that a white male with a degree is over 11 times more likely to be of an upper class income status than that of a non-white female with no degree:
```{r dataframe, echo = FALSE}
data.frame(
race = c( 1, 0, 1, 1, 0, 0, 1, 0),
sex = c( 1, 1, 0, 1, 0, 1, 0, 0),
degree = c( 1, 1, 1, 0, 1, 0, 0, 0),
p = c("57%", "48%", "27%", "24%", "18%", "16%", "8%", "5%")
)
```


---
#Conclusive Results
Our findings clearly indicate that following demographic has the largest proportion of upper-class citizens:

--

- Older age

--

- Male

--

- White

--

- Having a Degree

--

## A Final Word

What we have quantified in here is alarming and disturbing. However, one aspect that makes this discrimination so difficult to address is that the experiment, as well-designed as it is, cannot send us much signal about why this prejudice is occurring. It is only possible to say that discrimination is happening. Finding strong evidence of racism, sexism and ageism for individual cases is a persistent challenge in enforcing anti-discrimination laws.
