---
title: "Demographic Features & Income"
author: "by THE BEST TEAM: Tom Cummings,Jon Francis, Finn Mclachlan, Roman Davies & Scott Gowan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, include = FALSE}
library(tidyverse)
library(tidymodels)
# Add any other libraries here



```


```{r load-data}

adult_data <- read.csv("data/adult.data")
```

In excess, income inequalities perpetuate economic instabilities and slower growth rates, as well as social discontent and political polarisation. Indeed, income inequality is a well-established controversy in economics. Whilst in part, these disparities can be accredited to varying capacities of effort, talent and luck; they also pertain to underlying prejudices and differences in demographic characteristics. In this project, we aim to highlight income inequalities between different demographics in the U.S. using reliable data from the UCLA census database. The results of our data analysis, ultimately predicting income levels based on demographic features, confirm the existence of such income inequalities and provide justifications where appropriate. Importantly, this investigation has been carried out not to further confirmation biases on political doctrine, but provide impartial evidence on whether these disparities exist, and if so, whether it is fair to attribute them exclusively to discrimination. 

Data:

The "Census Income" or “adult.data” dataset is a collection of information related to individuals and their income from the UC Irvine Machine Learning Repository, derived from the 1994 U.S. census with over 33,000 entries. This includes age (from over 16 onwards), education, race, gender and hours worked per week as the variables that we have deemed most prominent in the discussion of income disparities and demographics; along with marital and relationship status, capital gain, capital loss, and native country, which will not feature. The dataset shows whether, based on these factors, an individual earns more than $50k annually. Granted this 1994 dataset may yield outdated results as cultural and economic factors change over time, though exaggerated in severity, these same trends of income inequality still stand today and so the data remains a feasible candidate for our exploration. 

Research Question:

How do various demographic factors contribute to income inequality, and can a predictive logistic model accurately identify these disparities within the 'Census Income' dataset? –
– Here, further clarifications must be made. Income levels have been binarised as either over or under-and-equal to $50k annually; categorised as ‘upper-class’ and ‘lower-class’ respectively. This investigation discerns inequalities solely based on the division of demographics into these two groups, simplifying our predictive model and allowing for more comprehensive interpretations of visualised data later on, lest overly complex financial figures. We also recognise that adult.data cannot alone paint the full picture. Many factors influence these disparities, and a full inspection would certainly require research and data that well surpasses the scope of this short report. Chiefly as a data science project, including additional data would not further demonstrate any of the skills acquired through this course; the methods behind our findings becoming overly repetitive. 

Methodology and Findings:

The framework for our approach is as follows: begin with data pre-processing to make less laborious the data interpretation; create visualisations to explore income distribution (in classes) across age, gender, education levels and race; construct a logistic regression model to predict income class against race, gender and qualifications; and finally, making a predictive model of income class for any given person based on their specific demographic features. We believe this procedure gives insight into the relationship between demographics and their income, as well as providing basic justifications for evident discrepancies.

Data Wrangling:

```{r}
#Changing the variable names for adult data
colnames(adult_data) <- c("age", "workclass", "id", "education", "education-num", "marital-status",
                          "occupation", "relationship", "race_old", "sex", "capital_gain",
                          "capital_loss", "hours_per_week", "native-country", "class")

```

First, after reading the two datasets above (adult.test including very similar statistics to adult.data), wrangling is required to prepare them for analysis. Initially our data did not contain column names, and so variable names were assigned. The aforementioned lesser important variables such as capital gain and loss were removed using the select() function from the ‘dplyr’ package. Next, new binary variables 'race' and 'has_degree' were created. 'Race' indicates whether a person is white or not; the 'is_white' function returns 1 if the original 'race_old' variable is "White", and 0 is returned otherwise. Mutating adds the new 'race' variable, and 'race_old' is removed in a similar fashion to capital gain and loss. Likewise, the 'has_degree' function will return 1 if the 'education' variable reads "Bachelors", "Masters" or "Doctorate", and 0 otherwise. The mutate() function adds the new 'has_degree' variable: (664)

```{r}
is_white <- function(x){ifelse(x == " White", 1, 0)}

adult_data <- adult_data %>% mutate(race = is_white(race_old))

has_degree <- function(x){ifelse(x == " Bachelors" | x == " Masters" | x == " Doctorate", 1, 0)}

adult_data <- adult_data %>% mutate(has_degree = has_degree(education))

adult_data <- adult_data %>% select(-capital_gain, -capital_loss, -race_old)

```

Age and Income Class:

```{r}
adult_data %>% ggplot() + geom_bar(mapping = aes(x = age, colour = class), position = "identity")

mean_age <- adult_data %>% group_by(class) %>% summarize(mean_age = mean(age))
mean_age
```
Age, though not through a prejudicial lens, is still an important determinant of income. We’ve created a bar plot using geom_bar() in ggplot2 to visualise the distribution of ages for the two income classes expressed in red and blue on the graph; the ‘Position = “identity”’ function allows the two curves to overlay. Additionally, we have calculated the mean age for each income class using the group_by and summarise functions, giving us a mean age for the upper-class as 44, and for the lower-class 37. This notable difference suggests that on average individuals in the upper-class are older than those in the lower-class. A potential contributing factor, captured by adult.data and scrutinised below, is the number of ‘hours-per-week’ worked:

```{r,echo=FALSE}
adult_data %>% ggplot() + geom_smooth(mapping = aes(x = age, y = hours_per_week))
```

```{r, echo=FALSE}

ggplot(data = adult_data, aes(x = age, y = hours_per_week)) +
  geom_point() +
  geom_smooth(method = "lm", 
              se = FALSE) +
  labs(
    title = "Age vs hours worked per week",
    x = "age (years)",
    y = "hours-per-week"
  )

linear_reg() %>%
  set_engine("lm") %>%
  fit(hours_per_week ~ age, data = adult_data) %>%
  tidy()

```

```{r, echo = FALSE}
hours_age_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(hours_per_week ~ age, data = adult_data)
  

hours_age_fit_aug <- augment(hours_age_fit$fit)
ggplot(hours_age_fit_aug, mapping = aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "gray", lty = "dashed") +
  labs(x = "Predicted hours worked", y = "Residuals")
```

Older people tend to work more hours and hence earn higher incomes. This is evidenced by the graphs above of ‘age’ versus ‘hours_per_week’, forming the linear regression equation of ‘hours__per_week = 38+0.0622 x age’. The coefficient for the age variable is 0.0622, indicating that as age increases by one year, the model predicts an additional 0.0622 hours worked per week. Upon evaluating a plot of residuals for the linear regression a subtle downwards trend emerges, albeit on a minute scale within the broader context of the plot. This observation suggests that the linear model may not be the optimal fit, with residuals exhibiting a slight deviation from the expected pattern. Further, our logistic regression model with ‘hours_per_week’ as the predictor shows that for every year a person ages, their probability of being predicted upper-class increases by 0.2%. Nevertheless, this relatively small positive association suggests that other factors are at play. upper-class individuals may have had more time to advance in their careers, pursue advanced education, accumulate wealth and specialise in their industry for higher earning potentials for example. Obviously, age alone may not be a direct cause of income disparities.

Gender and Income Class:

```{r, echo=FALSE}
percentage_sex_data <- adult_data %>%
  group_by(sex, class) %>%
  summarize(count = n()) %>%
  mutate(percentage = count / sum(count) * 100)

ggplot(percentage_sex_data, aes(x = sex, y = percentage, fill = class)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), vjust = -0.5, position = position_dodge(width = 0.9)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
The curious relationship between gender and income have led to heated debate over probable inherent workplace biases (i.e. the gender-pay-gap). Our graphed interpretation of the dataset confirms this discrepancy in male-female income. Within our code, ‘geom_bar(stat = "identity", position = "dodge")’ creates a grouped bar plot where each gender has adjacent bars, and the height of each bar represents the percentage of a specific income class within that gender. With 30.6%, males have around three times the quantity of women with 10.9% as their respective proportions in the upper-class.

``` {r,echo=FALSE}
adult_data %>% ggplot() + geom_density(mapping = aes(x = hours_per_week, colour = sex))

```

Again, one possible explanation for such a difference could be in contrasting hours worked. A new graph, overlaying the sexes and their ‘hours_per_week’ worked shows a discrepancy in mean weekly working hours, with men averaging 42.4 and women 36.4. The higher average for men may indicate a societal or occupational trend where men, on average, engage in professions requiring longer hours (some occupations also yield higher pay regardless of hours). Various factors, including societal expectations and caregiving responsibilities, especially in the 90s, may contribute to this difference. These mean values however, do not provide a nuanced understanding of individual circumstances, and it is likely that female income was, on average, subject to some discrimination. (1191)

Education and Income Class:

```{r,echo=FALSE}
education_counts <- adult_data %>%
  group_by(education) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

adult_data$education <- adult_data$education %>% factor(levels = education_counts$education)

adult_data %>% ggplot() + geom_bar(mapping = aes(x = education, fill = class), position = "dodge") + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

In a similar fashion to the gender-income bar plot, a graph of education level versus income class has been established. Certainly, significant impacts of qualifications on income are expected. Our code first calculates and arranges the count of observations for each education level, creating a data frame named ‘education_counts.’ Subsequently, it reorders the levels of  'education' based on these counts. Each bar in the plot represents observational counts for different income classes within each education level. While it may appear that 'HS-Grad' has the highest population in the lower-class and 'Bachelors' dominates the upper-class, it's important to note that these observations are influenced by the larger overall populations of these educational groups rather than indicating the inherent impact of education on income levels. A more meaningful analysis involves considering the percentage representation of each education level within their respective income classes (i.e. 'Doctorates' may proportionally lead to the highest representation in the upper-class, offering a more insightful perspective). Consequently, a new plot will be generated to illustrate these percentage-based results, correcting for the misrepresentation that raw counts may introduce and providing a clearer understanding of the predictive power of education levels in determining income levels.

```{r, echo = FALSE}
adult_data %>% ggplot() + geom_bar(mapping = aes(x = has_degree, fill = class), position = "dodge") + labs( x = "No Degree                           Degree")

#Note that 47% of people with degrees are upper-class whereas only 17% of people without degrees are upper-class
```

Utilising the 'has_degree' binary variable, we discern a notable concentration of upper-class individuals within the 'Degree' group, contrasting with a smaller portion in the 'No Degree' category. Yet, due to the dataset's predominant focus on individuals without degrees, a modest percentage still yields a higher count of upper-class individuals without degrees. This can be explained positively, levelling imbalanced educational opportunities.

Race and Income Class:

```{r, echo=FALSE}
adult_data %>% ggplot() + geom_bar(mapping = aes(x = race, fill = class), position = "dodge") + labs( x = "Not White                                                        White")

# 34% of white people are upper-class
# 18% of not white people are upper-class.


```

Exploring another demographic aspect, our code visually represents the distribution of income classes across diverse racial categories in the 'race' variable. The plot employs the dodge position to arrange bars side-by-side for different racial categories. Subsequently, a bar plot depicting the percentage of individuals by race and income class is created. Notably, when it comes to attaining upper-class status, minorities face less favourable odds in comparison to their counterparts. Specifically, the figures reveal that 34% of white individuals are eligible for considerations as upper-class, contrasting with 18% for all minorities. Justifications for this disparity were not convincingly conveyed through the dataset. Racist dogma continually influences employment and incomes, and hence it is plausible to suggest that more nuanced factors that affect income from race, such as occupational choices and education levels, may still stem from inherent prejudice. (1588)

Predictive model:

```{r}
adult_data$class <- factor(adult_data$class)

class_fit <- logistic_reg() %>%
  set_engine("glm") %>%
  fit(class ~ race + sex + has_degree, data = adult_data, family = "binomial")

tidy(class_fit)
```

Finally, we formed a predictive model built upon 'race,' 'sex,' and 'has_degree' as fundamental income differentiators. First calculating conditional probabilities for a new individual's income class based on specified demographic features, we provide a direct prediction. Then transforming the 'class' variable into a factor, fitting a logistic regression model using 'race,' 'sex,' and 'has_degree' as predictors, and displaying coefficients, elucidates how these features collectively influence income class likelihood. Moreover, using binarised variables simplifies and makes more accurate predictions. 

The logit function ‘log(1/(1-p))=-2.96 + (0.494*race) + (1.3*sex) + (1.45*degree)’ derived from the logistic regression model then expresses the log-odds of belonging to the upper-class. The intercept (-2.96) signifies the baseline log-odds, while positive coefficients for 'race' (0.494), 'sex' (1.3), and 'degree' (1.45) indicate increases in the likelihood of being in the upper-class. Exponentiating these coefficients yields odds ratios, conveying the multiplicative change in odds for a one-unit change in each predictor variable. 

```{r dataframe, echo = FALSE}
data.frame(
race = c( 1, 0, 1, 1, 0, 0, 1, 0),
sex = c( 1, 1, 0, 1, 0, 1, 0, 0),
degree = c( 1, 1, 1, 0, 1, 0, 0, 0),
p = c("57%", "48%", "27%", "24%", "18%", "16%", "8%", "5%")
)
```
Disconcertingly, our findings are indicative of large disparities of income in certain demographic groups. A white male with a degree is over 11x more likely to be upper-class than a non-white female without one. To re-emphasise, our results cannot fully quantify the reasons behind disparities, but only provide evidence for and speculate over their existence. Further, anti-discrimination laws are hard to implement with deep complexity around contributing factors. If any consolation, the socio-political landscape has greatly shifted since 1994, and at present, the progressive nature of income inequality is clear. 

##References:

Horesh, Yair, Noa Haas, Yehezkel Resheff, and Ron Kohavi. 1996. “Census Income.” UCI Machine Learning Repository. https://archive.ics.uci.edu/dataset/20/census+income.