---
title: "Project title"
author: "by Team-Name: User1, User2, User3, User4 & User5"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, include = FALSE}
library(tidyverse)
# Add any other libraries here



```


```{r load-data, include=FALSE}
# load your data here and any pre-processing/cleaning 
#that you need for your report
# setwd("~/IDS/GroupProject/data/census+income")
adult_data <- read.csv("adult.data")
adult_test <- read.csv("adult.csv")

colnames(adult_data) <- c("age", "workclass", "fnlwgt", "education", "education-num", "marital-status", "occupation", "relationship", "race", "sex", "capital-gain", "capital-loss", "hours-per-weed", "native-country", "class")





```
##notes
all ages are above 16

remember to make diagnostics plots for any linear models

take into account the type of people who will have taken part in the census
The data was taken in 1994 
 
we can make a prediction algorith to determines someones class based on all of the other data 

What we have quantified in this section is alarming and disturbing. However, one aspect that makes this racism so difficult to address is that the experiment, as well-designed as it is, cannot send us much signal about which employers are discriminating. It is only possible to say that discrimination is happening, even if we cannot say which particular callbacks — or non-callbacks — represent discrimination. Finding strong evidence of racism for individual cases is a persistent challenge in enforcing anti-discrimination laws.

## Research Question

Our aim is to look investigate the relationship between income and personal variables. This means looking at how factors such as age, race, gender, degree and looking at how their income class throughout western society. Our project hopes to make evident any clear discrepancies in income that may be as a result of person variables. We then hope to use this data to create a prediction model that will determine the likelihood of person being of upper class income, that is making over 50K a year. Furthermore we also carry out an analysis which attempts to explain the relationship between age and income. Our hypothesis was that older people tend to work higher hours and people who work higher hours generally make more money hence older people tend to make more money - that is of course up until the age of retirement.

Investigating variables that affect income from a data set of 33000 individuals from a UCLA census.


## Data

Our data set is called "Adult_data" which contains

Our main dataset is called "Adult data" which contains... and is from... 

we then also used a data set taken more recently in order to compare them...

A Census of 3300 people taken in 1994 that includes information on adults(over 16's): age, race, sex,native-country , class, etc.

archive.ics.uci.edu/dataset/20/census+income a 






## Findings


-Introduction:

By extracting data on over 32000 people from a census taken in 1994 by UCLA we aim to find out which variables, such as gender, race and age, have the largest affect on a persons income. Hence our research into the income inequality between identities should make clear whether or not their is any clear prejudice between certain types of people and if so which types of people are discriminated against the most. Furthermore we will then extrapolate this data using modeling methods in order to predict a persons class based on their identity.

Keep in mind that our research is not large enough scale to consider more than two gender's - however if we were to continue to research deeply into the topic then other sex's would most certainly be of utmost consideration.

Any form of discrimination is concerning, and this is why we decided it was so important to discuss this topic using data.The primary objective of our data science project is to develop a robust predictive logistic model capable of estimating an individual's income based on several key parameters: race, having a degree, gender, workclass, age, marital status, occupation, relationship status, capital gain, capital loss, hours per week, native country, and income (categorized as above or below 50k annually). This predictive model aims to provide an accurate estimation of income level by leveraging these key factors. In creating this model we hope to reveal which variables are the most detrimental to a persons income class hence we hope to make clear any discrimination.

For our investigation, we have selected the "adult" dataset from the UC Irvine Machine Learning Repository. This dataset comprises over 33 thousand entries and includes 13 distinct variables. We have chosen this dataset due to its credibility and richness in variables, expecting it to serve as a reliable foundation for our analysis and model development.

It's crucial to note that the data we're working with originates from a census conducted in 1994 and only considers adults over the age of 16. While this dataset offers substantial insights, the temporal and age gap might introduce challenges due to potential changes in societal, economic, and cultural factors over time. We will address and mitigate these potential limitations as part of our analysis process.

Through this report, we aim to explore, analyze, and build predictive models that effectively estimate an individual's income based on the provided variables. Our ultimate goal is to develop an accurate predictor while acknowledging and accounting for any potential biases or limitations introduced by the dataset's age.


-First steps:
Our first steps of the project were mainly looking at the data and getting a feel for how we would go about elaborating the predictor. We started off by doing some simple wrangling where we eliminated some NA values from the the dataset as well as eliminating some variables like capital gain and capital loss as it represents the amount of money won or lost in assets (mainly stocks and shares by each member , something we thought would not be important as the majority of people have a very small pool of assets. Also after looking at the data both of these columns had virtually all values of zero so we decided to be selective of unnecessary data such as this. Furthermore, we also changed the name of the column previously called fnlwgt to ID where we reached a point where we were happy with our data and had a clear and organised data set to start from.


-Graphical methods to visualize relationships between variables 
With our data wrangling complete we then decided to make a few graphs using the ggplot2 package to visualise the data and try to identify relationships between the variables. We started off plotting a barchart with race along the x axis. For every race we used two different bars, one for the percentage of people in that race with income over 50k (red) and one for the percentage of people in that race earning less or equal to 50k (bleu). We used these different colours as well as percentages on the y axis to successfully visualise the ratios of people in these races earning over 50k. We also went against using the staggered function as we believed showing two bars next to each other would be a better form of visualisation for anyone viewing our work. We then repeated the same process for gender as we also wanted to have a clear idea of how gender influences the propensity for a person to have incomes over 50k.

-conclusion

Our results very clearly point towards the idea that people who are white, male or have a university degree are far more likely to be in a higher income category. However the data analysis is limited to merely that. Yes, our data points towards this despicable prejudice however it cannot explain why this discrimination is happening. However we must also take into account the fact that this census was taken in 1994 hence advances in society may mean that the data may, hopefully, differ in the our current civilization 

Our data analysis makes clear that there is infact a clear discrimination however it does not explain wh, it merely proved its existance

we must laso take into account that the census was in 1994 fro which scoiety has oved fowar much better since then so the numbers may look different if the data was from a more recent census.

## References

archive.ics.uci.edu/dataset/20/census+income 
https://openintro-ims.netlify.app/model-logistic.html

